\documentclass[11pt]{article}
\usepackage{wasysym}
\usepackage{listings}
\usepackage{tikz}
\usepackage{url}
%\usepackage{algorithm2e}
\usetikzlibrary{arrows,automata,shapes}
\tikzstyle{block} = [rectangle, draw, fill=blue!20, 
    text width=2.5em, text centered, rounded corners, minimum height=2em]
\tikzstyle{bw} = [rectangle, draw, fill=blue!20, 
    text width=3.5em, text centered, rounded corners, minimum height=2em]

\newcommand{\handout}[5]{
  \noindent
  \begin{center}
  \framebox{
    \vbox{
      \hbox to 5.78in { {\bf ECE459: Programming for Performance } \hfill #2 }
      \vspace{4mm}
      \hbox to 5.78in { {\Large \hfill #5  \hfill} }
      \vspace{2mm}
      \hbox to 5.78in { {\em #3 \hfill #4} }
    }
  }
  \end{center}
  \vspace*{4mm}
}

\newcommand{\lecture}[4]{\handout{#1}{#2}{#3}{#4}{Lecture #1}}
\topmargin 0pt
\advance \topmargin by -\headheight
\advance \topmargin by -\headsep
\textheight 8.9in
\oddsidemargin 0pt
\evensidemargin \oddsidemargin
\marginparwidth 0.5in
\textwidth 6.5in

\parindent 0in
\parskip 1.5ex
%\renewcommand{\baselinestretch}{1.25}

\begin{document}
\lecture{4 --- January 12, 2015}{Winter 2015}{Patrick Lam}{version 1}

\section*{Multicore Processors}
As I've alluded to earlier, multicore processors came about because
clock speeds just aren't going up anymore. We'll discuss technical
details today.

Each processor \emph{core} executes instructions; a processor with
more than one core can therefore simultaneously execute multiple
(unrelated) instructions. 

\paragraph{Chips and cores.} Multiprocessor (usually SMP, or symmetric
multiprocessor) systems have been around for a while. Such systems
contain more than one CPU. We can count the number of CPUs by
physically looking at the board; each CPU is a discrete physical thing.

Cores, on the other hand, are harder to count. In fact, they look just
like distinct CPUs to the operating system:

{\scriptsize \begin{verbatim}
plam@plym:~/courses/p4p/lectures$ cat /proc/cpuinfo
processor	: 0
vendor_id	: GenuineIntel
cpu family	: 6
model		: 23
model name	: Pentium(R) Dual-Core  CPU      E6300  @ 2.80GHz
...
processor	: 1
vendor_id	: GenuineIntel
cpu family	: 6
model		: 23
model name	: Pentium(R) Dual-Core  CPU      E6300  @ 2.80GHz
\end{verbatim}
}
\vspace*{-1em}
%$ 
If you actually opened my computer, though, you'd only find one chip.
The chip is pretending to have two \emph{virtual CPUs}, and the
operating system can schedule work on each of these CPUs. In general,
you can't look at the chip and figure out how many cores it contains.

\vspace*{-1em}
\paragraph{Hardware Designs for Multicores.}
In terms of the hardware design, cores might share a cache, as in this picture:

\vspace*{-1em}
\begin{center}
\begin{tikzpicture}
\draw (-0.5, -0.5) rectangle ++(0.9, 0.9);
\draw (-0.05, 0.05) node {\small Core};

\draw (-0.5, -1.5) rectangle ++(0.9, 0.9);
\draw (-0.05, -1.05) node {\small Core};

\draw (0.5, -1.5) rectangle ++(1.2, 1.9);
\draw (1.1, -.55) node {\small Cache};

\draw (-0.6, -1.6) rectangle ++(2.45, 2.1);

\draw (2.2, -1.6) rectangle ++(2.7, 2.1);
\draw (3.6, -0.55) node {\small Rest of system};

\draw[<->] (1.9, -0.55) -- (2.15, -0.55);
\end{tikzpicture}
\end{center}
\vspace*{-1em}
\hfill (credit: \emph{Multicore Application Programming}, p. 5)

This above Symmetric Multithreading (SMP) design is especially good for the 1:1 threading model. In
this case, the design of the cores don't need to change much, but they
still need to communicate with each other and the rest of the system.

Or, we can have a design that works well for the N:1 model:
\begin{center}
\begin{tikzpicture}
\draw (-1.4, -1.5) rectangle ++(1.7, 1.9);
\draw (-0.55, 0.05) node {\small Thread 1};
\draw (-0.55, -1.05) node {\small Thread 2};
\draw (-0.55, -0.5) node {\small Core};

\draw (0.5, -1.5) rectangle ++(1.05, 1.9);
\draw (1.05, -.55) node {\small Cache};

\draw (-1.5, -1.6) rectangle ++(3.15, 2.1);

\draw (2.2, -1.6) rectangle ++(2.7, 2.1);
\draw (3.6, -0.55) node {\small Rest of system};

\draw[<->] (1.7, -0.55) -- (2.15, -0.55);
\end{tikzpicture}
\end{center}

One would expect that executing two threads on one core might
mean that each thread would run more slowly. It depends on the instruction
mix. If the threads are trying to access the same resource, then each
thread would run more slowly. If they're doing different things, there's 
potential for speedup.

Finally, it's possible to both use multiple cores and put multiple threads
onto one core, as in the M:N model:

\begin{center}
\begin{tikzpicture}

 \draw (-1.4, -1.5) rectangle +(1.7, 1.9);
 \draw (-1.4, -3.5) rectangle +(1.7, 1.9);

\foreach \y in {-2, 0}
{
 \draw (-0.55, 0.05) + (0, \y) node {\small Thread 1};
 \draw (-0.55, -1.05) + (0, \y) node {\small Thread 2};
 \draw (-0.55, -0.5) + (0, \y) node {\small Core};
}

\draw (0.5, -3.5) rectangle ++(1.05, 3.9);
\draw (1.05, -1.55) node {\small Cache};

\draw (-1.5, -3.6) rectangle ++(3.15, 4.1);

\draw (2.2, -3.6) rectangle ++(2.7, 4.1);
\draw (3.6, -1.55) node {\small Rest of system};

\draw[<->] (1.7, -1.55) -- (2.15, -1.55);
\end{tikzpicture}
\end{center}
Here we have four hardware threads; pairs of threads share hardware
resources. One example of a processor which supports chip multi-threading
(CMT) is the UltraSPARC T2, which has 8 cores, each of which supports 8 
threads. All of the cores share a common level 2 cache.

\paragraph{Non-SMP systems.} The designs we've seen above have been
more or less SMP designs; all of the cores are mostly alike.
A very non-Smp system is the Cell, which contains a PowerPC main core
(the PPE) and 7 Synergistic Processing Elements (SPEs), which are
small vector computers.


\paragraph{Non-Uniform Memory Access.}
In SMP systems, all CPUs have approximately the same access time for resources
(subject to cache misses). There are also NUMA, or Non-Uniform Memory Access,
systems out there. In that case, CPUs can access different resources at different
speeds. (Resources goes beyond just memory).

In this case, the operating system should schedule tasks on CPUs which
can access resources faster.  Since memory is commonly the bottleneck,
each CPU has its own memory bank.

\paragraph{Using CMT effectively.} Typically, a CPU will expose its
hardware threads using virtual CPUs.
In current hardware designs, each of the hardware threads has the same
performance. 

However, performance varies depending on context. In the above
example, two threads running on the same core will most probably run
more slowly than two threads running on separate cores, since they'd
contend for the same core's resources. Task switches between cores (or
CPUs!) are also slow, as they may involve reloading caches.

Solaris ``processor sets''
enable that operating system to assign processes to specific virtual CPUs,
while Linux's ``affinity'' keeps a process running on the same virtual CPU.
Both of these features reduce the number of task switches, and processor
sets can help reduce resource contention, along with Solaris's locality
groups\footnote{Gove suggests that locality groups help reduce contention
for core resources, but they seem to help more with memory.}

\section*{Processes versus Threads}
The first design decision that you need to solve when parallelizing programs
is whether you should use threads or processes.

  \begin{itemize}
    \item Threads are basically light-weight processes which piggy-back on
          processes' address space.
  \end{itemize}

Traditionally (pre-Linux 2.6) you had to use {\tt
  fork}~(for~processes) and {\tt clone} (for threads). But {\tt clone}
is not POSIX compliant, and its man page says that it's
Linux-specific---FreeBSD uses {\tt rfork()}. (POSIX is the standard
for Unix-like operating systems).

\paragraph{When processes are better.} {\tt fork} is safer and more secure than threads.

  \begin{enumerate}
    \item Each process has its own virtual address space:
      \begin{itemize}
        \item Memory pages are not copied, they are copy-on-write. Therefore, processes use less memory than you would expect.
      \end{itemize}
    \item Buffer overruns or other security holes do not expose other~processes.
    \item If a process crashes, the others can continue.
  \end{enumerate}

{\bf Example:} In the Chrome browser, each tab is a separate
process. Scott McCloud explained this:
\url{http://uncivilsociety.org/2008/09/google-chrome-comic-by-scott-m.html}.

\paragraph{When threads are better.} Threads are easier and faster.

  \begin{enumerate}
    \item Interprocess communication (IPC) is more complicated and slower than interthread
          communication; must use operating system utilities (pipes, semaphores, shared
              memory, etc) instead of thread library (or just memory reads and writes).
    \item Processes have much higher startup, shutdown and synchronization costs than threads.
    \item Pthreads fix the issues of {\tt clone} and provide a
          uniform interface for most systems. (You'll work with them in Assignment 1.)
  \end{enumerate}

\paragraph{How to choose?}
  If your application is like this:
  \begin{itemize}
    \item mostly independent tasks, with little or no communication;
    \item task startup and shutdown costs are negligible compared to overall runtime; and
    \item want to be safer against bugs and security holes,
  \end{itemize}
then processes are the way to go. If it's the opposite of this, then use threads.

For performance reasons, along with ease and consistency across systems, we'll use threads. We will describe both Pthreads and C++ 11 threads, in particular.

\end{document}
